---
title: "Lab 1: NYT API"
author: "Your Name"
date: "2023-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(jsonlite) #convert results from API queries into R-friendly formats 
library(tidyverse)
library(tidytext) #text data management and analysis
library(ggplot2) #plot word frequencies and publication dates
```


```{r setup, include=FALSE}
#assign API key.  When you create a NYT Dev account, you will be given a key
#API_KEY <- "YOUR_API_KEY"
 API_KEY <- "p1AVHKnC9rIXfgzuco6vj5eFgryns1gItCibpzuEbjGO5UYP"
```

Today we will be grabbing some data from the New York Times database via their API, then running some basic string manipulations, trying out the tidytext format, and creating some basic plots.

<https://developer.nytimes.com/>

### Connect to the New York Times API and send a query

We have to decide which New York Times articles we are interested in examining. For this exercise, we'll examine coverage of **extreme heat** — one of the most consequential and fast-accelerating climate hazards. Heat waves are now hotter, longer, and more frequent due to climate change, and they kill more people globally than any other weather-related disaster. 

<!-- Media coverage of extreme heat has also shifted dramatically: what was once framed as a weather story is increasingly treated as a climate and public health crisis. That makes it a rich topic for text analysis — we can examine not just *how much* coverage there is, but *how* the language and framing evolves over time. -->

We'll send a query to the NY Times API using a URL that contains information about the articles we'd like to access.

`fromJSON()` is a wrapper function that handles our request and the API response. We'll use it to create an object `initialhits` with the results of our query. The `flatten = TRUE` argument converts from the nested JSON format to an R-friendlier form.

```{r api, eval = FALSE}
searchQ = URLencode("extreme heat")
url = paste0('https://api.nytimes.com/svc/search/v2/articlesearch.json?q=', searchQ,
             '&begin_date=20240401&end_date=20260401&api-key=', API_KEY)

t = fromJSON(url, flatten = TRUE)
hits = t$response$meta$hits
print(hits)
```

```{r api, eval = FALSE}
#what type of object is t?
class(t) 

t <- data.frame(t)

# how big is it?
dim(t)

# what variables are we working with?
names(t)

```

Let's look at a piece of text. Our data frame has a column called `snippet` that contains a short excerpt, or "snippet" from the article. Let's grab a snippet and try out some basic `stringr` functions.

The `stringr` package gives us a consistent set of functions for working with character strings. All functions start with `str_` and take the string as the first argument, making them easy to use inside `mutate()` or `filter()` calls. Before we tokenize the full corpus, it's worth knowing these tools — they'll come up constantly in text cleaning.

```{r basic_stringr, eval=FALSE}
#assign a snippet to x to use as fodder for stringr functions.  You can follow along using the sentence on the next line.
snip <- dat$snippet[9]

x <- "Governments around the world are enacting measures to try to protect workers from the dangers of heat stress. They’re barely keeping up with the risks."

tolower(x)
str_split(x, ','); str_split(x, 't')
str_replace(x, 'dangers', 'threats')
str_replace(x, ' ', '_') #first one
#how do we replace all of them?
str_replace_all(x,' ', '_')
str_detect(x, 't'); str_detect(x, 'heat') ### is pattern in the string? T/F
str_locate(x, 't'); str_locate_all(x, 'heat')
```


### OK, it's working but we want more data. Let's set some parameters for a bigger query.


```{r api-function}
nytAPI = function(term1, begin_date, end_date, api) {
  searchQ = URLencode(term1)
  url = paste0('https://api.nytimes.com/svc/search/v2/articlesearch.json?q=', searchQ,
               '&begin_date=', begin_date, '&end_date=', end_date, '&api-key=', api)
  
  initialsearch = fromJSON(url, flatten = TRUE) #we create an initial search to find out how many total hits, or articles that meet our query parameters
  hits = initialsearch$response$meta$hits
  
  if (hits == 0) return(data.frame()) ##if there are no hits, return an empty data frame

  maxPages = min(round(hits / 10) - 1, 10) #otherwise, we find out how many times we'll need to loop thru fcn, each time returning 10 pages

  df = data.frame(id=as.numeric(), created_time=character(),
                  snippet=character(), headline=character())

  for (i in 0:maxPages) {
    nytSearch = fromJSON(paste0(url, "&page=", i), flatten = TRUE)
    temp = data.frame(id = 1:nrow(nytSearch$response$docs),
                  created_time = nytSearch$response$docs$pub_date,
                  snippet = nytSearch$response$docs$snippet,
                  headline = nytSearch$response$docs$headline.main)
    df = rbind(df, temp)
    Sys.sleep(12)
  }
  return(df)
}
```

```{r results}
results = nytAPI(term1 = "extreme heat", begin_date = "20240401", end_date = "20260401", api = API_KEY)

write.csv(results, paste0("NYT-", term1,".csv"))
```

<!-- The initial query returned one page of ten articles, but also gave us count of total number of hits on our query. We can use that to size a for() loop to automate requests. -->

<!-- ```{r #hits, eval=FALSE} -->
<!-- #dig into the JSON object to find total hits -->
<!-- initialQuery <- fromJSON(baseurl) -->
<!-- maxPages <- round((initialQuery$response$meta$hits[1] / 10)-1)  -->
<!-- #initiate a list to hold results of our for loop -->
<!-- pages <- list() -->

<!-- #loop -->
<!-- for(i in 0:maxPages){ -->
<!--   nytSearch <- fromJSON(paste0(baseurl, "&page=", i), flatten = TRUE) %>% data.frame()  -->
<!--   message("Retrieving page ", i) -->
<!--   pages[[i+1]] <- nytSearch  -->
<!--   Sys.sleep(20)  -->
<!-- } -->
<!-- ``` -->

<!-- We converted each returned JSON object into a data frame -->

<!-- ```{r, bind_rows} -->
<!-- class(nytSearch) -->
<!-- class(pages) -->
<!-- #need to bind the pages and create a tibble from nytDat -->
<!-- df <- do.call("rbind", pages) -->

<!-- saveRDS(nyt_df, "nyt_df.rds") -->
<!-- ``` -->

Load the preconstructed nytDat so you can follow along.

```{r date-plot}
nytDat <- read_csv("/courses/EDS232/Data/covtype_sample.csv")
dim(nytDat)
df <- nytDat
```

```{r}
nytDat %>%
  mutate(pubDay = as.Date(created_time)) %>%
  group_by(pubDay) %>%
  summarise(count=n()) %>%
  filter(count >= 2) %>%
  ggplot() +
  geom_bar(aes(x=reorder(pubDay, count), y=count), stat="identity") +
  coord_flip()
```

The New York Times doesn’t make full text of the articles available through the API. But we can use the first paragraph of each article.

### Tokenization

**Tokenization** is the process of splitting text into individual units — usually words, but sometimes phrases or characters. It’s the first step in almost every text analysis workflow because most NLP methods operate on tokens, not raw strings. The `tidytext` function `unnest_tokens()` handles tokenization and simultaneously converts to lowercase, strips punctuation, and puts the data in tidy (one-token-per-row) format, which plays well with `dplyr` and `ggplot2`.

```{r plot_frequencies}
names(nytDat)

paragraph <- names(nytDat)[4] #The 6th column, "response.doc.lead_paragraph", is the one we want here.

#use tidytext::unnest_tokens to put in tidy form
tokenized <- nytDat %>%
  unnest_tokens(word, paragraph) #word is the new column, paragraph is the source

tokenized[,"word"]
```

Now we can count word frequencies across all first paragraphs. This raw frequency distribution gives us a first look at what dominates the corpus — though as you’ll see, the top results aren’t very informative yet.

```{r word_frequencies}
tokenized %>%
  count(word, sort = TRUE) %>%
  filter(n > 100) %>% #illegible with all the words displayed
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word)) +
  geom_col() +
  labs(y = NULL)
```

Uh oh, who knows what we need to do here?

### Stop Word Removal

The most frequent words in any corpus are almost always **stop words** — function words like "the", "a", "of", "and" that carry little semantic content on their own. Removing them lets the meaningful, content-bearing words rise to the top of our frequency counts. The `tidytext` package ships with a `stop_words` lexicon drawn from three dictionaries (SMART, Snowball, onix) that covers the most common English stop words. We remove them with `anti_join()`, which keeps only rows in our token data that do *not* appear in the stop words list.

```{r stop-words}

data(stop_words)
stop_words

tokenized <- tokenized %>%
  anti_join(stop_words)

tokenized %>%
  count(word, sort = TRUE) %>%
  filter(n > 10) %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word)) +
  geom_col() +
  labs(y = NULL)
```

```{r inspect-tokens}
#inspect the list of tokens (words)
tokenized$word
```

### Text Normalization and Cleaning

Even after removing stop words, raw tokens often contain noise: digits mixed into words, possessives like `"heat’s"`, stray punctuation, or empty strings left behind after other substitutions. **Text normalization** (sometimes called regularization) is the process of standardizing token form so that variants of the same surface string are treated consistently. Here we use `stringr` and base R regex to strip digits and possessives, then remove any empty strings that result.

```{r cleaning, eval=FALSE}
clean_tokens <- str_remove_all(tokenized$word, "[:digit:]") #remove all numbers

clean_tokens <- gsub("’s", '', clean_tokens) 

tokenized$clean <- clean_tokens

tokenized %>%
  count(clean, sort = TRUE) %>%
  mutate(clean = reorder(clean, n)) %>%
  filter(n > 5) %>%
  ggplot(aes(n, clean)) +
  geom_col() +
  labs(y = NULL)
```

## Assignment (Due Tuesday 4/11 11:59pm)

1.  Create a free New York Times account (<https://developer.nytimes.com/get-started>)

2.  Pick an interesting environmental key word(s) and use the {jsonlite} package to query the API. Pick something high profile enough and over a large enough time frame that your query yields enough articles for an interesting examination.

3.  Recreate the publications per day and word frequency plots using the snippets.  

-   Make some (at least 3) transformations to the corpus (add context-specific stopword(s), stem a key term and its variants, remove numbers)

4.  Recreate the publications per day and word frequency plots using the headlines variable (response.docs.headline.main). Compare the distributions of word frequencies between the first paragraph and headlines. Do you see any difference?
