---
title: "Lab 1: NYT API"
author: "Mateo Robbins"
date: "2024-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(jsonlite) #convert results from API queries into R-friendly formats 
library(tidyverse)
library(tidytext) #text data management and analysis
library(ggplot2) #plot word frequencies and publication dates

#assign API key.  When you create a NYT Dev account, you will be given a key
API_KEY <- "NTKBHbsb6XFEkGymGumAiba7n3uBvs8V"
```

Today we will be grabbing some data from the New York Times database via their API, then running some basic string manipulations, trying out the tidytext format, and creating some basic plots.

<https://developer.nytimes.com/>

### Connect to the New York Times API and send a query

We have to decide which New York Times articles we are interested in examining. For this exercise, I chose articles about Deb Haaland, the current US Secretary of the Interior. As a member of the Laguna Pueblo Tribe, Haaland is the first Native American to serve as a Cabinet secretary. Very cool!

We'll send a query to the NY Times API using a URL that contains information about the articles we'd like to access.

fromJSON() is a wrapper function that handles our request and the API response. We'll use it to create an object,t, with the results of our query. The flatten = T argument converts from the nested JSON format to an R-friendlier form.

```{r api, eval = FALSE}

#create the query url
url <- paste("http://api.nytimes.com/svc/search/v2/articlesearch.json?q=haaland&api-key=",API_KEY, sep ="")

#send the request, receive the response, and flatten
t <- fromJSON(url, flatten = T)
```

```{r api, eval = FALSE}
#what type of object is it?
class(t) 

#convert to a data frame 
t <- data.frame(t)

# how big is it?
dim(t)

# what variables are we working with?
names(t)

```

The name format, "response.xxx.xxx...", is a legacy of the JSON nested hierarchy.

Let's look at a piece of text. Our data object has a variable called "response.docs.snippet" that contains a short excerpt, or "snippet" from the article. Let's grab a snippet and try out some basic string manipulations from {stringr}.

```{r basic_stringr, eval=FALSE}
t$response.docs.snippet[9]

#assign a snippet to x to use as fodder for stringr functions.  You can follow along using the sentence on the next line.

x <- "Her nomination as secretary of the interior is historic, but as the first Native cabinet member, she would have to strike a delicate balance." 

tolower(x) #convert to lowercase.  Standard text cleaning step
str_split(x, ','); str_split(x, 't') #split into substrings, returns a list.  
str_replace(x, 'historic', 'without precedent') #swap one string for another
str_replace(x, ' ', '_') #str_replace replaces first occurrence of pattern. 
#how do we replace all of them?
str_replace_all(x,' ', '_')
#note: gsub() from base R is commonly used and essentially equivalent to str_replace_all()
str_detect(x, 't'); str_detect(x, 'tive') ### str_detect lets us ask if a pattern is in the string? Boolean T/F. Good for filtering or applying conditional operations
str_locate(x, 't'); str_locate_all(x, 'as') #we can also ask where its occuring. Start and end positions
```

### OK, it's working but we want more data. Let's set some parameters for a bigger query.

```{r}
term1 <- "Deb" 
term2 <- "&Haaland" # Need to use $ to string  together separate terms
begin_date <- "20210120"
end_date <- "20240401"

#construct the query url using API operators
baseurl <- paste0("http://api.nytimes.com/svc/search/v2/articlesearch.json?q=",term1,term2,"&begin_date=",begin_date,"&end_date=",end_date,"&facet_filter=true&api-key=",API_KEY, sep="")

#examine our query url
baseurl
```

The initial query returned one page of ten articles, but also gave us count of total number of hits on our query. We can use that to size a for() loop to automate requests.

```{r, eval=FALSE}
#dig into the JSON object to find total hits
initialQuery <- fromJSON(baseurl)
#Can dig in to find max results, but I'll shorten it to 10 for demo
#maxPages <- round((initialQuery$response$meta$hits[1] / 10)-1) 
maxPages <- 10
#initiate a list to hold results of our for loop
pages <- list()

## partially complete loop
for(i in 0:maxPages){
  nytSearch <- fromJSON(paste0(baseurl, "&page=", i), flatten = TRUE) %>% data.frame() 
  message("Retrieving page ", i)
  pages[[i+1]] <- nytSearch 
  Sys.sleep(12) 
}
```

We converted each returned JSON object into a data frame.

```{r, bind_rows}
class(nytSearch)
class(pages)
#need to bind the pages and create a tibble from nytDat
nyt_df <- bind_rows(pages)

```

Let's start exploring our data.  What types of content did we turn up?
```{r article-type}
#response.docs 3 times
nyt_df %>% 
  group_by(response.docs.type_of_material) %>%
  summarize(count=n()) %>% #This creates a new data frame with the count of records for each type_of_material.
  mutate(percent = (count / sum(count))*100) %>% #add percent of total column
  ggplot() +
  geom_bar(aes(y=percent, x=response.docs.type_of_material, fill=response.docs.type_of_material), stat = "identity") + coord_flip()
```

```{r date-plot}
nytDat

nytDat %>%
  mutate(pubDay=gsub("T.*","",response.docs.pub_date)) %>% 
  group_by(pubDay) %>%
  summarise(count=n()) %>%
  filter(count >= 2) %>%
  ggplot() +
  geom_bar(aes(x=reorder(pubDay, count), y=count), stat="identity") +
  coord_flip() #bring date so bars go longwise
```

The New York Times doesn't make full text of the articles available through the API. But we can use the first paragraph of each article.

```{r plot_frequencies}
names(nytDat)
head(nytDat[,1:6])
nytDat[6] #The 6th column, "response.doc.lead_paragraph", is the one we want here. 

#use tidytext::unnest_tokens to put in tidy form.  
#If there are some types of news that we'd like to exclude, we can filter()
tokenized <- nytDat %>%
  filter(response.docs.news_desk!="Sports") %>%
unnest_tokens(word, response.docs.lead_paragraph) #word is the new column, paragraph is the source

tokenized[,"word"]
```

Alright, let's starting analyzing our data.  What the most frequent words in the articles we have?
```{r word_frequencies}
tokenized %>%
  count(word, sort = TRUE) %>%
  filter(n > 100) %>% #illegible with all the words displayed
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word)) +
  geom_col() +
  labs(y = NULL)
```

Uh oh, who knows what we need to do here?

We'll load a dictionary of stop words, or high frequency, low information words, which we would like to exclude from our data.

```{r stop-words}

data(stop_words) #load stop_words from {tidytext}
stop_words # a 2-column table.  The first contains individual words, the second the names of a few different stop word lexicons combined here.

#we'll do an anti_join to remove words from out data that occur in a stop word lexicon.
tokenized <- tokenized %>%
  anti_join(stop_words)

#now let's try that plot again
tokenized %>%
  count(word, sort = TRUE) %>%
  filter(n > 10) %>% #this time use a much lower filter
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word)) +
  geom_col() +
  labs(y = NULL)
```

OK, but look at the most common words. Does anything stand out?

```{r cleaning, eval=FALSE}

#inspect the list of tokens (words)
tokenized$word

clean_tokens <- str_remove_all(tokenized$word, "[:digit:]") #remove all numbers

clean_tokens <- gsub("’s", '', clean_tokens)

tokenized$clean <- clean_tokens

tokenized %>%
  count(clean, sort = TRUE) %>%
 #illegible with all the words displayed
  mutate(clean = reorder(clean, n)) %>%
  ggplot(aes(n, clean)) +
  geom_col() +
  labs(y = NULL)

#remove the empty strings
tib <-subset(tokenized, clean!="")

#reassign
tokenized <- tib

#try again
tokenized %>%
  count(clean, sort = TRUE) %>%
  filter(n > 10) %>% 
  mutate(clean = reorder(clean, n)) %>%
  ggplot(aes(n, clean)) +
  geom_col() +
  labs(y = NULL)

```

## Assignment (Due Tuesday 4/9 11:59pm)
Reminder: Please suppress all long and extraneous output from your submissions (ex:  lists of tokens).

1.  Create a free New York Times account (<https://developer.nytimes.com/get-started>)

2.  Pick an interesting environmental key word(s) and use the {jsonlite} package to query the API. Pick something high profile enough and over a large enough time frame that your query yields enough articles for an interesting examination.

3.  Recreate the publications per day and word frequency plots using the first paragraph field.  This time filter on the response.docs.news_desk variable to winnow out irrelevant results.

-   Make some (at least 3) transformations to the corpus including: add context-specific stopword(s), stem a key term and its variants, remove numbers)

4.  Recreate the publications per day and word frequency plots using the headlines variable (response.docs.headline.main). Compare the distributions of word frequencies between the first paragraph and headlines. Do you see any difference?
