---
title: "Lab 2"
author: "Your Name"
date: "Date"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(LexisNexisTools)
library(dplyr)
library(readr)
library(stringr)
library(here)
library(tidytext)
library(tidyr) #pivot_wider()
library(ggplot2)
```

Sentiment analysis is a tool for assessing the mood of a piece of text. For example, we can use sentiment analysis to understand public perceptions of topics in environmental policy like energy, climate, and conservation.

### Introduction to the example

Today's example will use data from the Nexis Uni (formerly Lexis Nexis) database, accessed through the UCSB library. There are a large number of news and other full-text publications available through the database. I'm interested in the discussion around deforestation of the Amazon rain forest. The current president of Brazil, Luiz Inacio Lula da Silva recently completed his first 100 days in office after running on a pledge to reduce the rate of deforestation. I'd like to know how the sentiment in coverage of deforestation changed after his election.

```{r raw_data}
setwd(here("Nexis/Post"))

my_files <- list.files(pattern = ".docx", path = getwd(),
                       full.names = TRUE, recursive = TRUE, ignore.case = TRUE)

```

We'll use the {LexisNexisTools} package to handle the documents from our Nexis search.

```{r}
```

```{r get_bing}
bing_sent <- get_sentiments('bing') #grab the bing sentiment lexicon from tidytext
head(bing_sent, n = 20)
```

1.  Score words using bing lexicon

```{r text_words}

```

2.  Calculate mean sentiment (by word polarity) across articles

```{r mean_sent}

```

3.  Sentiment by article plot

Let's try a very basic plot to see the amount of positive vs. negative articles.

```{r plot_sent_scores}

```

##### Origin of the NRC lexicon

"These guys selected about 10,000 words from an existing thesaurus... and then created a set of five questions to ask about each word that would reveal the emotions and polarity associated with it. That's a total of over 50,000 questions.

They then asked these questions to over 2000 people, or Turkers, on Amazon's Mechanical Turk website, paying 4 cents for each set of properly answered questions.

The result is a comprehensive word-emotion lexicon for over 10,000 words."

4.  nrc emotion words

Let's take a look at the most common emotion words in the data set

```{r nrc_sentiment}

```

Let's break it out and plot the contributions by particular emotion categories.

```{r sent_counts}

```

Now let's do a quick comparison to articles from the 100-days leading up to the beginning of Lula's term.

```{r}
setwd(here("Nexis/Pre"))

pre_files <- list.files(pattern = ".docx", path = getwd(),
                       full.names = TRUE, recursive = TRUE, ignore.case = TRUE)

pre_dat <- lnt_read(pre_files)

pre_meta_df <- pre_dat@meta
pre_articles_df <- pre_dat@articles
pre_paragraphs_df <- pre_dat@paragraphs

pre_dat2<- tibble(Date = pre_meta_df$Date, Headline = pre_meta_df$Headline, id = pre_dat@articles$ID, text = pre_dat@articles$Article)
```

```{r pre_text_words}
text_words <- pre_dat2  %>%
  unnest_tokens(output = word, input = text, token = 'words')
 
 sent_words <- text_words%>% #break text into individual words
  anti_join(stop_words, by = 'word') %>% #returns only the rows without stop words
  inner_join(bing_sent, by = 'word') #joins and retains only sentiment words
```

```{r mean_pre}
pre_sentiment <- sent_words %>%
  count(id, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>% 
  mutate(polarity = positive - negative)
```

### Assignment (Due 4/18 by 11:59 PM)

1.  Access the Nexis Uni database through the UCSB library: <https://www.library.ucsb.edu/research/db/211>
2.  Choose a key search term or terms to define a set of articles.
3.  Use your search term along with appropriate filters to obtain and download a batch of at least 100 full text search results (.docx).

-   Guidance for {LexisNexisTools} : <https://github.com/JBGruber/LexisNexisTools/wiki/Downloading-Files-From-Nexis>

4.  Read your Nexis article document into RStudio.
5.  This time use the full text of the articles for the analysis. First clean any artifacts of the data collection process (hint: this type of thing should be removed: "Apr 04, 2022( Biofuels Digest: <http://www.biofuelsdigest.com/> Delivered by Newstex"))
6.  Explore your data a bit and replicate the analyses above presented in class.
7.  Look at the nrc contribution to emotion by word plots. Identify and reclassify or remove at least one term that gives misleading results in your context.
8.  Plot the amount of nrc emotion words as a percentage of all the emotion words used each day (aggregate text from articles published on the same day). How does the distribution of emotion words change over time? Can you think of any reason this would be the case?
