---
title: 'Lab 3: Topic Analysis'
output:
  word_document: default
  pdf_document: default
---

```{r packages}
library(quanteda)
library(tm)
library(topicmodels)
library(ldatuning)
library(tidyverse)
library(tidytext)
library(reshape2)

```

Load the data

```{r data}
tbl <-read_csv("https://raw.githubusercontent.com/MaRo406/EDS-231-text-sentiment/main/data/sb_dat_text.csv")
```

First we'll build the corpus using corpus() from {quanteda}.

```{r corpus}

```

Next, we can use tokens(), also from {quanteda}, to construct a tokens object. tokens() takes a range of arguments related to cleaning the data. Next we'll create a stopwords lexicon and remove each word contained in it from our tokens object. The quanteda function tokens_select() lets us do the removal.

```{r tokens}
```

Now we can transform our data to prepare for topic modeling. Let's create a document-feature matrix with quanteda::dfm(). Topic modeling doesn't work with empty rows in this matrix, so we'll need to remove those. I do that here using {slam}, which is designed to deal with sparse matrices like ours.

```{r dfm}
```

Great, now we are almost ready to run a model. We just have to come up with an initial value for k, the number of latent topics present in the data. How do we do this? There are multiple methods. I think there are three main ideas you (well, ChatGPT) wrote about in your stories: weather, travel, and relaxation. So I will tell the model to look for 3 topics by setting the k parameter = 3.

```{r LDA_modeling}

```

Running topicmodels::LDA() produces two posterior probability distributions: theta, a distribution over k topics within each document and beta,the distribution v terms within each topic, where v is our vocabulary (total unique words in our data set).

Let's examine at our results. posterior() extracts theta and beta

```{r LDA_modeling}

```

Alright, so that worked out OK. An alternative to specifying k based on theory or a hypothesis is to run a series of models using a range of k values. ldatuning::FindTopicsNumber gives us the tools for this.

```{r find_k}

```

Alright, now let's estimate another model, this time with our new value of k.

```{r LDA_again}
k <- 5

topicModel_k5 <- LDA(dfm, k, method="Gibbs", control=list(iter = 500, verbose = 25))

tmResult <- posterior(topicModel_k5)
terms(topicModel_k5, 10)
theta <- tmResult$topics
beta <- tmResult$terms
vocab <- (colnames(beta))

```

There are multiple proposed methods for how to measure the best k value. You can [go down the rabbit hole here](https://rpubs.com/siri/ldatuning)

```{r top_terms_topic}

```

```{r plot_top_terms}

top_terms %>%
  mutate(term = reorder_within(term, beta, topic, sep = "")) %>%
  ggplot(aes(term, beta, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free_y") +
  scale_x_reordered()+
  coord_flip()

```

Let's assign names to the topics so we know what we are working with. We can name the topics by interpreting the overall theme or idea they represent, but for now let's just name them by their top terms.

```{r topic_names}

```

We can explore the theta matrix, which contains the distribution of each topic over each document.

```{r topic_dists}

# get topic proportions from example documents
 
ggplot(data = viz_df, aes(topic, value, fill = document), ylab = "proportion") +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +  
  coord_flip() +
  facet_wrap(~ document, ncol = N)
```

Here's a neat JSON-based model visualizer, {LDAviz}.  We can use this to visualize the words-on-topics distribution and intertopic distances.  The size of the circles in the LDAvis plot show proportionally the amount of words that belong to each topic, and the space between circles shows the degree to which the circles share words.

```{r LDAvis}
library(LDAvis) #visualization 
library("tsne") #matrix decomposition
svd_tsne <- function(x) tsne(svd(x)$u)
json <- createJSON(
  phi = tmResult$terms, 
  theta = tmResult$topics, 
  doc.length = rowSums(dfm), 
  vocab = colnames(dfm), 
  term.frequency = colSums(dfm),
  mds.method = svd_tsne,
  plot.opts = list(xlab="", ylab="")
)
serVis(json)

```
The relevance param. defines the relationship between the word's topic probability and its lift. Lift is the ratio of the probability of the word in the topic to the probability of the word in the corpus. The more frequent is the word in the corpus, the lower will be its lift. Saliency also gives a idea of both how common the word is in a topic to overall how common it is.  Same vein as tf-idf

### Assignment Lab 3:

Due in 2 weeks: May 2 at 11:59PM

For this assignment you'll the articles data you downloaded from Nexis Uni in Week 2.

1.  Create a corpus from your articles.

2.  Clean the data as appropriate.

3.  Run three models (i.e. with 3 values of k) and select the overall best value for k (the number of topics) - include some justification for your selection: theory, FindTopicsNumber() optimization metrics, interpretability, LDAvis

4. Plot the top terms in each topic and the distribution of topics across a sample of the documents (constrained by what looks good in the plot).

5.  Take a stab at interpreting the resulting topics. What are the key themes discussed in the articles in your data base?
