---
title: "Lab 4"
author: "Mateo Robbins"
date: "2023-04-25"
output: html_document
---

```{r packages, include = FALSE}
library(tidytext)
library(tidyverse)
library(tidymodels)
library(textrecipes)
library(discrim) # naive-bayes
```

This data set includes more possible predictors than the text alone, but for this model we will only use the text variable
```{r data}

urlfile ="https://raw.githubusercontent.com/MaRo406/EDS-231-text-sentiment/main/data/climbing_reports_model_dat.csv"
incidents_df<-readr::read_csv(url(urlfile))
```

Now we'll split our data into training and test portions

```{r split-data}
set.seed(1234)

incidents_split 

incidents_train 
incidents_test 

```

We use recipe() to specify the predictor and outcome variables and the data.

```{r recipe}
```

Next we add some familiar pre-processing steps on our Text variable: tokenize to word level, filter to the most common words, and calculate tf-idf.

```{r pre-process}
```

Create  tidymodels workflow to combine the modeling components

```{r workflow}

```

```{r nb-spec}

```

Now we are ready to add our model to the workflow and fit it to the training data

```{r fit-model}

```
Next up is model evaluation. We'll stretch our training data further and use resampling to evaluate our naive Bayes model. Here we create 10-fold cross-validation sets, and use them to estimate performance.

```{r cv_folds}

```

```{r nb-workflow}

```

To estimate its performance, we fit the model to each of the resampled folds, and then evaluate on the heldout part of each resampled fold.

```{r fit-resamples}

```

Extract the relevant information using collect_metrics() and collect_predictions() and examine the performance metrics.

```{r performance}

```

We'll use two performance metrics: accuracy and ROC AUC.
Accuracy is the proportion of the data that is predicted correctly. 
The ROC curve plots the true positive rate against the false positive rate; AUC closer to 1 indicates a better-performing model, while AUC closer to 0.5 indicates a model that does no better than random guessing.

```{r performance-plot}

```

Another model method involves the confusion matrix. A confusion matrix tabulates a model's false positives and false negatives for each class.

```{r confusion-matrix}

```

```{r null-model}

```
