---
title: "Week 5: Topic Modeling Lab Preparation"
subtitle: "EDS 231 — Spring 2026"
date: "April 21, 2026"
output:
  html_document:
    theme: bootstrap
    highlight: pygments
    css: week-styles.css
    toc: false
    toc_float:
      collapsed: false
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> Selection of articles to serve as data for next week's topic modeling lab.

---

### Learning Objectives

By the end of NEXT week, you will be able to:

- Explain the generative assumptions behind Latent Dirichlet Allocation (LDA)
- Construct a document-feature matrix (DFM) suitable for topic modeling using `{quanteda}`
- Select a value of k using theory, optimization metrics (`ldatuning`), and interpretability
- Interpret the beta (word-topic) and theta (topic-document) posterior distributions
- Visualize topic structure with term plots and `{LDAvis}`
- Apply topic modeling to a real news corpus and characterize emergent themes

---

### Recommended Reading

📖 [Text Mining with R — Ch. 6: Topic Modeling](https://www.tidytextmining.com/topicmodeling)

📖 [Blei, Ng & Jordan (2003) — Latent Dirichlet Allocation, *JMLR*](https://jmlr.org/papers/v3/blei03a.html)

---

## Lecture Materials

### Slides

🖥 Slide deck from lecture

[View Slides](https://ucsb.box.com/s/qge5x1j6gtwh1gd6j0hywc4vv8gjhb8p)

### Code & Data

💻 R scripts used during class:

[]()

---

## Preparing for Lab 4: Getting Your Data from Nexis Uni

Lab 4 will ask you to run a topic model on a corpus of news articles **you retrieve yourself** from Nexis Uni. Follow the steps below before lab so you arrive with data ready to go.

### Step 1 — Search Nexis Uni

1. Go to the [UCSB Library Nexis Uni portal](https://www.library.ucsb.edu/) and search for **Nexis Uni**.
2. Run a keyword search on a topic of your choosing.
3. Filter to **News** sources. You can use any other filters you deem appropriate for your purposes. Aim for around **100** articles** — enough for meaningful topics but not so many that model fitting is slow.
4. Select download format of**Word (.docx)**. You'll have to select articles in batches for download.

### Step 2 — Organize your files

Place all downloaded `.docx` files in a folder inside your project:

```
your-project/
  Nexis/
    your-topic/
      export1.docx
      export2.docx
```

### Step 3 — Read the data into R

The `{LexisNexisTools}` package parses the Nexis `.docx` exports. Install it if you haven't already:

```r
install.packages("LexisNexisTools")
```

Then use the code below to load your files and assemble a tidy data frame. The commented-out block below is the standard workflow — update the `path` argument to match your folder name:

```{r load-nexis, eval=FALSE}
library(LexisNexisTools)
library(tidyverse)

pre_files <- list.files(
  path = here::here("Nexis", "your-topic"),   # update to your folder
  pattern = "\\.docx$",
  full.names = TRUE,
  recursive = TRUE,
  ignore.case = TRUE
)

pre_dat <- lnt_read(pre_files)

pre_meta_df       <- pre_dat@meta
pre_articles_df   <- pre_dat@articles
pre_paragraphs_df <- pre_dat@paragraphs

df <- tibble(
  date     = pre_meta_df$Date,
  headline = pre_meta_df$Headline,
  id       = pre_meta_df$ID,
  text     = pre_articles_df$Article,
  source   = pre_meta_df$Newspaper
)
```
---

## 📂 Assignment not due next week

### Lab 4 — Topic Modeling with Nexis Uni Data <span class="badge badge-danger">Graded · 11 pts</span>

**Due: May 12 @ 11:59 pm**
---

## 📚 Topic Modeling in Environmental Science — Key Citations

Curated readings organized by thematic area. Useful background for contextualizing your Lab 4 corpus and interpreting your results.

<details>
<summary><strong>Foundational Methods</strong></summary>
<table>
<thead><tr><th>Citation</th><th>Year</th><th>Topic</th><th>Keywords</th></tr></thead>
<tbody>
<tr><td><a href="https://www.jmlr.org/papers/v3/blei03a.html">Blei, Ng &amp; Jordan — <em>J. Machine Learning Research</em></a></td><td>2003</td><td>LDA Methodology</td><td>LDA, probabilistic topic model, generative model, Bayesian inference</td></tr>
<tr><td><a href="https://dl.acm.org/doi/10.1007/s11042-018-6894-4">Jelodar et al. — <em>Multimedia Tools &amp; Applications</em></a></td><td>2019</td><td>LDA Survey</td><td>LDA survey, Gibbs sampling, variational inference, model extensions</td></tr>
<tr><td><a href="https://pmc.ncbi.nlm.nih.gov/articles/PMC9120935/">Egger &amp; Yu — <em>Frontiers in Artificial Intelligence</em></a></td><td>2022</td><td>Social Media Topic Modeling</td><td>LDA, NMF, BERTopic, short-text, model comparison</td></tr>
</tbody>
</table>
</details>

<details>
<summary><strong>Environmental Science Research Trends</strong></summary>
<table>
<thead><tr><th>Citation</th><th>Year</th><th>Topic</th><th>Keywords</th></tr></thead>
<tbody>
<tr><td><a href="https://www.sciencedirect.com/science/article/abs/pii/S0959652621025373">Palanichamy &amp; Kargar — <em>Journal of Cleaner Production</em></a></td><td>2021</td><td>ESE Research Trends</td><td>LDA, environmental science, research trends, temporal analysis</td></tr>
<tr><td><a href="https://pmc.ncbi.nlm.nih.gov/articles/PMC7312947/">Ma, Li &amp; Zhang — <em>IJERPH</em></a></td><td>2020</td><td>Bibliometrics</td><td>highly cited papers, climate change, microplastics, Web of Science</td></tr>
<tr><td><a href="https://link.springer.com/article/10.1007/s10640-021-00554-0">Bjørner — <em>Environmental &amp; Resource Economics</em></a></td><td>2022</td><td>Environmental Economics</td><td>environmental economics, climate change, energy economics, trend detection</td></tr>
</tbody>
</table>
</details>

<details>
<summary><strong>Biodiversity Conservation</strong></summary>
<table>
<thead><tr><th>Citation</th><th>Year</th><th>Topic</th><th>Keywords</th></tr></thead>
<tbody>
<tr><td><a href="https://www.researchgate.net/publication/281059725_Text_analysis_tools_for_identification_of_emerging_topics_and_research_gaps_in_conservation_science">Westgate et al. — <em>Conservation Biology</em></a></td><td>2015</td><td>Conservation Research Gaps</td><td>topic modeling, emerging topics, research gaps, text analysis</td></tr>
<tr><td><a href="https://pmc.ncbi.nlm.nih.gov/articles/PMC8525079/">Nolan et al. — <em>Methods in Ecology &amp; Evolution</em></a></td><td>2021</td><td>Conservation Biology Trends</td><td>topic modeling, biodiversity hotspots, NLP, abstract analysis</td></tr>
<tr><td><a href="https://pmc.ncbi.nlm.nih.gov/articles/PMC4240467/">Valle et al. — <em>Ecology Letters</em></a></td><td>2014</td><td>Biodiversity Data Analysis</td><td>LDA, species assemblages, community ecology, forest succession</td></tr>
<tr><td><a href="https://pmc.ncbi.nlm.nih.gov/articles/PMC11959320/">Edelmann et al. — <em>Conservation Biology</em></a></td><td>2025</td><td>NLP for Evidence Synthesis</td><td>NLP, LLMs, evidence synthesis, conservation social science</td></tr>
<tr><td><a href="https://conbio.onlinelibrary.wiley.com/doi/10.1111/csp2.12670">IUCN Red List Threat Analysis — <em>Conservation Science &amp; Practice</em></a></td><td>2022</td><td>Threat Identification</td><td>IUCN Red List, extinction threats, habitat destruction, threat prioritization</td></tr>
</tbody>
</table>
</details>

<details>
<summary><strong>Corporate Environmental Reporting</strong></summary>
<table>
<thead><tr><th>Citation</th><th>Year</th><th>Topic</th><th>Keywords</th></tr></thead>
<tbody>
<tr><td><a href="https://pmc.ncbi.nlm.nih.gov/articles/PMC5389611/">Székely &amp; vom Brocke — <em>PLOS ONE</em></a></td><td>2017</td><td>Sustainability Reports</td><td>LDA, GRI, sustainability reporting, CSR, longitudinal analysis</td></tr>
<tr><td><a href="https://www.sciencedirect.com/science/article/pii/S105752192400601X">Müllerleile et al. — <em>ScienceDirect</em></a></td><td>2024</td><td>ESG Reporting Trends</td><td>BERTopic, ESG, annual reports, climate risk, Non-Financial Reporting Directive</td></tr>
<tr><td><a href="https://pmc.ncbi.nlm.nih.gov/articles/PMC9275432/">Park, Choi &amp; Jung — <em>Frontiers in Psychology</em></a></td><td>2022</td><td>ESG Public Discourse</td><td>LDA, Dynamic Topic Model, ESG, Twitter, sentiment analysis</td></tr>
<tr><td><a href="https://www.sciencedirect.com/science/article/pii/S0167923624001027">Kriebel &amp; Foege — <em>Decision Support Systems</em></a></td><td>2024</td><td>NLP Method Comparison</td><td>NLP, BERT, ChatGPT, sustainability disclosure, 10-K filings</td></tr>
<tr><td><a href="https://www.mdpi.com/2071-1050/18/3/1486">Gorovaia et al. — <em>Sustainability</em></a></td><td>2025</td><td>Greenwashing Detection</td><td>greenwashing, LDA, TF-IDF, ESG disclosure, corporate transparency</td></tr>
</tbody>
</table>
</details>

<details>
<summary><strong>Pollution Monitoring from Social Media</strong></summary>
<table>
<thead><tr><th>Citation</th><th>Year</th><th>Topic</th><th>Keywords</th></tr></thead>
<tbody>
<tr><td><a href="https://pmc.ncbi.nlm.nih.gov/articles/PMC4400579/">Feng et al. — <em>PLOS ONE</em></a></td><td>2015</td><td>Air Quality / Weibo</td><td>LDA, Weibo, PM2.5, air quality, social media sensing</td></tr>
<tr><td><a href="https://www.sciencedirect.com/science/article/abs/pii/S0959652621009884">Wang &amp; Jia — <em>Journal of Cleaner Production</em></a></td><td>2021</td><td>Social Media → Policy</td><td>Weibo, PM2.5, environmental governance, bottom-up pressure</td></tr>
<tr><td><a href="https://www.sciencedirect.com/science/article/pii/S1574954124001924">Yao et al. — <em>Ecological Informatics</em></a></td><td>2024</td><td>Contaminated Sites</td><td>topic modeling, public perception, risk management, remediation</td></tr>
<tr><td><a href="https://pmc.ncbi.nlm.nih.gov/articles/PMC8156109/">Lin et al. — <em>IJERPH</em></a></td><td>2021</td><td>Air Pollution Adaptation</td><td>LDA, air pollution, social media mining, adaptation behavior</td></tr>
<tr><td><a href="https://www.sciencedirect.com/science/article/abs/pii/S0301479723020984">Chang et al. — <em>Journal of Environmental Management</em></a></td><td>2023</td><td>Urban Pollution Complaints</td><td>Twitter, civil complaints, urban pollution, spatial overlap</td></tr>
</tbody>
</table>
</details>

---
