---
title: 'Lab 4: Topic Analysis'
author: "Mateo Robbins"
date: 
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Background: Latent Dirichlet Allocation

Latent Dirichlet Allocation (LDA) is a generative probabilistic model
that treats each document as a mixture of topics, and each topic as a
mixture of words. Understanding the key variables helps interpret model
outputs:

| Symbol | Name | Role |
|-----------------------------|----------------------|----------------------|
| **α** (alpha) | Hyperparameter | Dirichlet prior on document-topic distributions. Small α → each document focuses on few topics. Set by researcher. |
| **β** (beta) | Hyperparameter | Dirichlet prior on topic-word distributions. Small β → each topic uses a focused vocabulary. Set by researcher. |
| **θ** (theta) | Document-topic distribution | Probability distribution over topics for each document. *P(topic \| document)*. Latent — must be inferred. |
| **φ** (phi) | Topic-word distribution | Probability distribution over vocabulary for each topic. *P(word \| topic)*. Called `beta` in tidytext. Latent — must be inferred. |
| **z** | Topic assignment | The hidden topic label allocated to each individual word token. The bridge between θ and φ. Latent — must be inferred. |
| **w** | Observed word | The only variable you actually see. Everything else is inferred from patterns in w. |

### The Generative Assumptions

LDA imagines that documents were written as follows:

1.  For each topic *t*, draw a word distribution **φ_t** \~ Dirichlet(β)
2.  For each document *m*, draw a topic distribution **θ_m** \~
    Dirichlet(α)\
3.  For each word position in that document:
    -   Draw a topic assignment **z** \~ Multinomial(θ_m)
    -   Draw the word **w** \~ Multinomial(φ_z)

Inference runs this process in reverse: given the observed words **w**,
recover the most probable **z**, **θ**, and **φ**.

### The Dirichlet Distribution

The Dirichlet distribution is a *distribution over distributions* — it
produces probability vectors that sum to 1, making it ideal for
representing topic mixtures and word mixtures. The concentration
parameter (α or β) controls sparsity:

-   **Small value (\< 1)**: samples are sparse — most probability sits
    on one or two categories. Realistic for most text corpora.
-   **Large value (\> 1)**: samples are spread evenly across all
    categories.
-   **Value = 1**: uniform — all mixtures equally likely.

------------------------------------------------------------------------

## Lab Setup

```{r packages}
library(quanteda)
library(tm)
library(topicmodels)
library(topicdoc)   
library(tidyverse)
library(tidytext)
library(reshape2)
library(tictoc)
library(LDAvis)
library(tsne)
library(servr)
```

## Load the Data

```{r data}
tbl <- read_csv("https://raw.githubusercontent.com/MaRo406/EDS-231-text-sentiment/refs/heads/main/data/nexis-dat.csv")
```

## Build the Corpus

We use `corpus()` from {quanteda} to build the corpus from our data
frame.

```{r corpus}
corpus <- 
```

## Tokenize and Clean

`tokens()` handles punctuation and number removal. We also build a
custom stopword list that removes common journalistic filler words that
carry no thematic signal. We can return to this section to add
additional words once we see the results.

```{r tokens}
add_stops <- 

toks <- 

toks1 <- 
```

## Build the Document-Feature Matrix

Topic modeling requires a document-feature matrix (DFM). We trim rare
terms (appearing in fewer than 2 documents) and remove empty rows — LDA
cannot process documents with no terms.

```{r dfm}
dfm1 <- 
dfm2 <- 

# Remove empty rows — LDA will error on these

```

------------------------------------------------------------------------

## Step 1: Initial Model (Theory-Driven k)

Great, now we are almost ready to run a model. We just have to come up
with an initial value for k, the number of latent topics present in the
data. How do we do this? Let's say I think there may be political,
economic and environmental articles. So I will tell the model to look
for 3 topics by setting the k parameter = 3.

```{r LDA_k3}
k <- 

topicModel_kXX <- 
```

### Inspect the Posterior Distributions

`posterior()` extracts the two key output matrices:

-   **theta (θ)**: document × topic matrix — *P(topic \| document)*
-   **beta/phi (φ)**: topic × term matrix — *P(term \| topic)*

```{r inspect_k3}
result <- posterior(topicModel_k6)

beta  <- 
theta <- 

cat("phi dimensions  (topics × vocab):", dim(beta), "\n")
cat("theta dimensions (docs × topics):", dim(theta), "\n")

# Top 10 terms per topic

```

------------------------------------------------------------------------

## Step 2: Selecting k with Coherence Scores

An alternative to theory-driven k selection is to fit models across a
range of k values and evaluate them statistically. We use
`topicdoc::topic_coherence()`, which measures whether the top words of
each topic co-occur frequently in the corpus — a proxy for human
interpretability. We can do this efficiently by running a sequence of
models, each with a different value of k.

**Higher coherence = more semantically meaningful topics.**

```{r find_k}


# Fit models for a range of k values
k_values <- 

coherence_results <- 



# Plot coherence vs k — look for the peak or elbow
ggplot(XXXXXXX, aes(x = , y = )) +
  geom_line(color = "navy") +
  geom_point(color = "navy", size = 2) +
  geom_vline(xintercept = coherence_results$k[which.max(coherence_results$coherence)],
             linetype = "dashed", color = "red") +
  labs(title = "Mean topic coherence by number of topics (k)",
       subtitle = "Dashed line = k with highest coherence",
       x = "Number of topics (k)",
       y = "Mean coherence score") +
  theme_minimal()
```

> **Interpreting the plot:** Look for the peak coherence value, or an
> "elbow" where adding more topics stops improving coherence
> meaningfully. Also inspect the top terms at two or three candidate k
> values — statistical criteria and human interpretability do not always
> agree. Your final choice of k should be justified on both grounds.

------------------------------------------------------------------------

## Step 3: Fit the Selected Model

Based on the coherence plot (and domain knowledge of the corpus), we
select a final k value and fit the final model.

```{r LDA_final}
k <-   # update this based on your coherence plot

topicModel_k_select <- LDA(dfm, k,
                           method = "Gibbs",
                           control = list(iter = 1000,
                                          verbose = 25,
                                          seed = 321))

tmResult <- posterior(topicModel_k_select)
terms(topicModel_k_select, 10)

theta <- 
beta  <-
vocab <-
```

------------------------------------------------------------------------

## Step 4: Visualize Top Terms per Topic

The β (φ) matrix gives us *P(term \| topic)*. Plotting the
highest-probability terms for each topic is the primary tool for human
interpretation of what each topic "means."

```{r top_terms_plot}
topics <- tidy(topicModel_k_select, matrix = "beta")

top_terms <- topics |>
  group_by(topic) |>
  slice_max(beta, n = 10) |>
  ungroup() |>
  arrange(topic, -beta)

top_terms |>
  mutate(term = reorder_within(term, beta, topic, sep = "")) |>
  ggplot(aes(term, beta, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free_y") +
  scale_x_reordered() +
  coord_flip() +
  labs(title = "Top 10 terms per topic",
       subtitle = "β = P(term | topic)",
       x = NULL, y = "β") +
  theme_minimal()
```

------------------------------------------------------------------------

## Step 5: Name the Topics

Interpret each topic by examining its top terms and assigning a
descriptive label. This is an inherently qualitative step — topic names
should reflect genuine thematic content, not just the top terms
mechanically concatenated.

```{r topic_names}
topic_words <- terms(topicModel_k_select, 5)

# Option 1: auto-name from top 5 terms (ok for a first pass)
topic_names <- apply(topic_words, 2, paste, collapse = " ")

# Option 2: replace with your own interpretive labels after inspection
# topic_names <- c("Topic Name 1", "Topic Name 2",
#                  "Topic Name 3",  "Topic Name 4"...)

topic_names
```

------------------------------------------------------------------------

## Step 6: Document-Topic Distributions (θ)

The θ matrix shows how each document blends across topics. This reflects
the allocation of word tokens **z** to topics within each document — a
document with many words assigned to topic 2 will have high θ for topic
2.

```{r topic_dists}
example_ids <- c(1:5)
n <- length(example_ids)

example_props <- theta[example_ids, ]
colnames(example_props) <- topic_names

viz_df <- melt(cbind(data.frame(example_props),
                     document = factor(1:n)),
               variable.name = "topic",
               id.vars = "document")

ggplot(data = viz_df, aes(topic, value, fill = document)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip() +
  facet_wrap(~ document, ncol = n) +
  labs(title = "Topic distributions across sample documents",
       subtitle = "θ = P(topic | document)",
       x = NULL, y = "Topic proportion")
```

------------------------------------------------------------------------

## Step 7: Interactive Visualization with LDAvis

LDAvis provides an interactive view of two complementary aspects of the
model:

-   **Intertopic distance map**: circles represent topics; size =
    proportion of corpus tokens assigned to that topic; distance =
    degree of topic overlap in vocabulary
-   **Term relevance bar chart**: adjustable by the λ (lambda) parameter

### The λ (lambda) Relevance Parameter

λ controls the trade-off between term frequency and term exclusivity
when ranking words within a topic:

$$\text{Relevance}(w, t) = \lambda \times P(w|t) + (1 - \lambda) \times \frac{P(w|t)}{P(w)}$$

-   **λ near 1**: ranks terms by raw frequency within the topic —
    highlights common words
-   **λ near 0**: ranks terms by how *exclusive* they are to the topic
    relative to the whole corpus — highlights distinctive words,
    analogous to tf-idf
-   **λ ≈ 0.6**: the empirically recommended default (Sievert & Shirley,
    2014) that balances both

```{r LDAvis}
svd_tsne <- function(x) tsne(svd(x)$u)

json <- createJSON(
  phi           = ,
  theta         = ,
  doc.length    = ,
  vocab         = ,
  term.frequency = ,
  mds.method    = svd_tsne,
  plot.opts     = list(xlab = "", ylab = "")
)

serVis(json)
```

> **Reading the LDAvis plot:** - Circles that are far apart share few
> words — good topic separation - Overlapping circles suggest those
> topics may be too similar and k could be reduced - Try λ = 0.6 as a
> starting point, then adjust to see distinctive vs. frequent terms

------------------------------------------------------------------------

## Assignment: Lab 4

For this assignment you will apply the full LDA pipeline to the Nexis
Uni article data you downloaded in Week 2. Create a new .Rmd and
complete the following:

**1. Build and clean your corpus**\
Create a corpus from your articles. Clean appropriately for your data —
consider what stopwords, minimum document frequencies, and tokenization
choices make sense given your specific corpus.

**2. Select k using multiple criteria**

Run three models (i.e. with 3 values of k) and select the overall best
value for k (the number of topics) - include some justification for your
selection: FindTopicsNumber() optimization metrics, theory,
interpretability, LDAvis. Select the best single value of k. Criteria to
consider:

\- **Optimization Metrics** - Run models across a range of k values and
produce a coherence plot using `topicdoc`.

\- **Theory**: what themes do you expect given your article topic?

\- **Interpretability**: read the top 10 terms at candidate k values —
do they form coherent, distinct themes?

\- **LDAvis**: do the topic circles separate well, or do they overlap
heavily?

Select a single best k and write 2–3 sentences justifying your choice
drawing on the three criteria.

**3. Fit your final model and visualize top terms**\
Plot the top 10 terms per topic.

**4. Visualize document-topic distributions**\
Plot θ for a sample of documents (5–10). Choose a sample size that
produces a readable plot.

**5. Interpret your topics**\
Assign a descriptive name to each topic (not just the concatenated top
terms). In a short paragraph per topic, describe:

\- What theme or issue does this topic represent?

\- Which words are most diagnostic of this interpretation?

\- Are there any surprising or ambiguous words in the top terms?
