---
title: 'Lab 4: Topic Analysis'
author: "Mateo Robbins"
date: 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Background: Latent Dirichlet Allocation

Latent Dirichlet Allocation (LDA) is a generative probabilistic model that treats each document as a mixture of topics, and each topic as a mixture of words. Understanding the key variables helps interpret model outputs:

| Symbol | Name | Role |
|--------|------|------|
| **╬▒** (alpha) | Hyperparameter | Dirichlet prior on document-topic distributions. Small ╬▒ тЖТ each document focuses on few topics. Set by researcher. |
| **╬▓** (beta) | Hyperparameter | Dirichlet prior on topic-word distributions. Small ╬▓ тЖТ each topic uses a focused vocabulary. Set by researcher. |
| **╬╕** (theta) | Document-topic distribution | Probability distribution over topics for each document. *P(topic \| document)*. Latent тАФ must be inferred. |
| **╧Ж** (phi) | Topic-word distribution | Probability distribution over vocabulary for each topic. *P(word \| topic)*. Called `beta` in tidytext. Latent тАФ must be inferred. |
| **z** | Topic assignment | The hidden topic label allocated to each individual word token. The bridge between ╬╕ and ╧Ж. Latent тАФ must be inferred. |
| **w** | Observed word | The only variable you actually see. Everything else is inferred from patterns in w. |

### The Generative Assumptions

LDA imagines that documents were written as follows:

1. For each topic *t*, draw a word distribution **╧Ж_t** ~ Dirichlet(╬▓)
2. For each document *m*, draw a topic distribution **╬╕_m** ~ Dirichlet(╬▒)  
3. For each word position in that document:
   - Draw a topic assignment **z** ~ Multinomial(╬╕_m)
   - Draw the word **w** ~ Multinomial(╧Ж_z)

Inference runs this process in reverse: given the observed words **w**, recover the most probable **z**, **╬╕**, and **╧Ж**.

### The Dirichlet Distribution

The Dirichlet distribution is a *distribution over distributions* тАФ it produces probability vectors that sum to 1, making it ideal for representing topic mixtures and word mixtures. The concentration parameter (╬▒ or ╬▓) controls sparsity:

- **Small value (< 1)**: samples are sparse тАФ most probability sits on one or two categories. Realistic for most text corpora.
- **Large value (> 1)**: samples are spread evenly across all categories.
- **Value = 1**: uniform тАФ all mixtures equally likely.

---

## Lab Setup

```{r packages}
library(quanteda)      # corpus, tokenization, and DFM construction
library(tm)            # legacy text-mining utilities; needed by some topicmodels internals
library(topicmodels)   # LDA() тАФ the core topic model fitting function
library(topicdoc)      # topic_coherence() тАФ statistical model evaluation
library(tidyverse)     # data wrangling and ggplot2 visualization
library(tidytext)      # tidy() method for LDA objects; reorder_within() for faceted plots
library(reshape2)      # melt() тАФ reshapes wide matrices to long format for ggplot
library(tictoc)        # tic() / toc() тАФ lightweight benchmarking: measures wall-clock time
library(LDAvis)        # createJSON() + serVis() тАФ interactive topic visualization
library(tsne)          # t-SNE dimensionality reduction used inside the LDAvis layout
library(servr)         # serves the LDAvis HTML in your browser via a local web server
```

## Load the Data

```{r data}
tbl <- read_csv("https://raw.githubusercontent.com/MaRo406/EDS-231-text-sentiment/refs/heads/main/data/nexis-dat.csv")
```

## Build the Corpus

We use `corpus()` from {quanteda} to build the corpus from our data frame.

```{r corpus}
corpus <- corpus(x = tbl, text_field = "text")
```

## Tokenize and Clean

`tokens()` handles punctuation and number removal. We also build a custom stopword list that removes common journalistic filler words that carry no thematic signal.

```{r tokens}
add_stops <- c(quanteda::stopwords("en", source = "smart"),
               "said", "says", "told", "according")

toks <- tokens(corpus, remove_punct = TRUE, remove_numbers = TRUE) |>
  tokens_tolower()

toks1 <- tokens_select(toks, pattern = add_stops, selection = "remove")
```

## Build the Document-Feature Matrix

Topic modeling requires a document-feature matrix (DFM). We trim rare terms (appearing in fewer than 2 documents) and remove empty rows тАФ LDA cannot process documents with no terms.

```{r dfm}
# dfm() converts the token list into a Document-Feature Matrix for LDA
dfm1 <- dfm(toks1)

dfm2 <- dfm_trim(dfm1, min_docfreq = 2)

# slam::row_sums() computes row totals on the sparse Matrix class used by quanteda.
sel_idx <- slam::row_sums(dfm2) > 0
dfm <- dfm2[sel_idx, ]
```

---

## Step 1: Initial Model (Theory-Driven k)

Great, now we are almost ready to run a model. We just have to come up with an initial value for k, the number of latent topics present in the data. How do we do this? Let's say I think there may be political, economic and environmental articles. So I will tell the model to look for 3 topics by setting the k parameter = 3.


```{r LDA_k3}
k <- 6

topicModel_k6 <- LDA(dfm,
                     k,
                     method = "Gibbs", #iterative resampling of topic labels for each token
                     control = list(iter = 1000,
                                    verbose = 25,
                                    seed = 321))  # set seed for reproducibility
```

### Inspect the Posterior Distributions

`posterior()` extracts the two key output matrices:

- **theta (╬╕)**: document ├Ч topic matrix тАФ *P(topic | document)*
- **beta/phi (╧Ж)**: topic ├Ч term matrix тАФ *P(term | topic)*

```{r inspect_k3}
result <- posterior(topicModel_k6)

beta  <- result$terms   # ╧Ж: topic-word distributions (rows = topics, cols = vocab)
theta <- result$topics  # ╬╕: document-topic distributions (rows = docs, cols = topics)

cat("phi dimensions  (topics ├Ч vocab):", dim(beta), "\n")
cat("theta dimensions (docs ├Ч topics):", dim(theta), "\n")

# Top 10 terms per topic
terms(topicModel_k6, 10)
```

---

## Step 2: Selecting k with Coherence Scores

An alternative to theory-driven k selection is to fit models across a range of k values and evaluate them statistically. We use `topicdoc::topic_coherence()`, which measures whether the top words of each topic co-occur frequently in the corpus тАФ a proxy for human interpretability.  We can do this efficiently by running a sequence of models, each with a different value of k.

**Higher coherence = more semantically meaningful topics.**


```{r find_k}
tic()

# Fit models for a range of k values
k_values <- seq(2, 15, by = 1) 

coherence_results <- map_dfr(k_values, function(k) { 
  m <- LDA(dfm, k = k, method = "Gibbs",
           control = list(iter = 300, seed = 321, verbose = 0))
  coh <- mean(topic_coherence(m, dfm))
  tibble(k = k, coherence = coh)
})

toc() #~185s 

# Plot coherence vs k тАФ look for the peak or elbow
ggplot(coherence_results, aes(x = k, y = coherence)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue", size = 2) +
  geom_vline(xintercept = coherence_results$k[which.max(coherence_results$coherence)],
             linetype = "dashed", color = "coral") +
  labs(title = "Mean topic coherence by number of topics (k)",
       subtitle = "Dashed line = k with highest coherence",
       x = "Number of topics (k)",
       y = "Mean coherence score") +
  theme_minimal()
```

> **Interpreting the plot:** Look for the peak coherence value, or an "elbow" 
> where adding more topics stops improving coherence meaningfully. 
> Also inspect the top terms at two or three candidate k values тАФ 
> statistical criteria and human interpretability do not always agree.
> Your final choice of k should be justified on both grounds.

---

## Step 3: Fit the Selected Model

Based on the coherence plot (and domain knowledge of the corpus), we select a final k value and fit the final model. 

```{r LDA_final}
k <- 6   # update this based on your coherence plot

topicModel_k_select <- LDA(dfm, k,
                           method = "Gibbs",
                           control = list(iter = 1000,
                                          verbose = 25,
                                          seed = 321))

tmResult <- posterior(topicModel_k_select)
terms(topicModel_k_select, 10)

theta <- tmResult$topics   # ╬╕: P(topic | document)
beta  <- tmResult$terms    # ╧Ж: P(term | topic)
vocab <- colnames(beta)
```

---

## Step 4: Visualize Top Terms per Topic

The ╬▓ (╧Ж) matrix gives us *P(term | topic)*. Plotting the highest-probability terms for each topic is the primary tool for human interpretation of what each topic "means."

```{r top_terms_plot}
topics <- tidy(topicModel_k_select, matrix = "beta")

top_terms <- topics |>
  group_by(topic) |>
  slice_max(beta, n = 10) |>
  ungroup() |>
  arrange(topic, -beta)

top_terms |>
  mutate(term = reorder_within(term, beta, topic, sep = "")) |>
  ggplot(aes(term, beta, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free_y") +
  scale_x_reordered() +
  coord_flip() +
  labs(title = "Top 10 terms per topic",
       subtitle = "╬▓ = P(term | topic)",
       x = NULL, y = "╬▓") +
  theme_minimal()
```

---

## Step 5: Name the Topics

Interpret each topic by examining its top terms and assigning a descriptive label. This is an inherently qualitative step тАФ topic names should reflect genuine thematic content, not just the top terms mechanically concatenated.

```{r topic_names}
topic_words <- terms(topicModel_k_select, 5)

# Option 1: auto-name from top 5 terms (quick but shallow)
topic_names <- apply(topic_words, 2, paste, collapse = " ")

# Option 2: replace with your own interpretive labels after inspection (recommended)
# topic_names <- c("Industrial pollution", "Conservation policy",
#                  "Corporate reporting",  "Climate adaptation")

topic_names
```

---

## Step 6: Document-Topic Distributions (╬╕)

The ╬╕ matrix shows how each document blends across topics. This reflects the allocation of word tokens **z** to topics within each document тАФ a document with many words assigned to topic 2 will have high ╬╕ for topic 2.

```{r topic_dists}
example_ids <- c(1:5)
n <- length(example_ids)

example_props <- theta[example_ids, ]
colnames(example_props) <- topic_names

viz_df <- melt(cbind(data.frame(example_props),
                     document = factor(1:n)),
               variable.name = "topic",
               id.vars = "document")

ggplot(data = viz_df, aes(topic, value, fill = document)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip() +
  facet_wrap(~ document, ncol = n) +
  labs(title = "Topic distributions across sample documents",
       subtitle = "╬╕ = P(topic | document)",
       x = NULL, y = "Topic proportion")
```

---

## Step 7: Interactive Visualization with LDAvis

LDAvis provides an interactive view of two complementary aspects of the model:

- **Intertopic distance map**: circles represent topics; size = proportion of corpus tokens assigned to that topic; distance = degree of topic overlap in vocabulary
- **Term relevance bar chart**: adjustable by the ╬╗ (lambda) parameter

### The ╬╗ (lambda) Relevance Parameter

╬╗ controls the trade-off between term frequency and term exclusivity when ranking words within a topic:

$$\text{Relevance}(w, t) = \lambda \times P(w|t) + (1 - \lambda) \times \frac{P(w|t)}{P(w)}$$

- **╬╗ near 1**: ranks terms by raw frequency within the topic тАФ highlights common words
- **╬╗ near 0**: ranks terms by how *exclusive* they are to the topic relative to the whole corpus тАФ highlights distinctive words, analogous to tf-idf
- **╬╗ тЙИ 0.6**: the empirically recommended default (Sievert & Shirley, 2014) that balances both

```{r LDAvis}
svd_tsne <- function(x) tsne(svd(x)$u)

json <- createJSON(
  phi           = tmResult$terms,
  theta         = tmResult$topics,
  doc.length    = rowSums(dfm),
  vocab         = colnames(dfm),
  term.frequency = colSums(dfm),
  mds.method    = svd_tsne,
  plot.opts     = list(xlab = "", ylab = "")
)

serVis(json)
```

---

## Assignment: Lab 4

For this assignment you will apply the full LDA pipeline to the Nexis Uni article data you downloaded in Week 2.

**1. Build and clean your corpus**  
Create a corpus from your articles. Clean appropriately for your data тАФ consider what stopwords, minimum document frequencies, and tokenization choices make sense given your specific corpus.

**2. Select k using multiple criteria**  

Run three models (i.e. with 3 values of k) and select the overall best value for k (the number of topics) - include some justification for your selection: theory, FindTopicsNumber() optimization metrics, interpretability, LDAvis. Select the best single value of k.

Run models across a range of k values and produce a coherence plot using `topicdoc`. Also consider:
- **Theory**: what themes do you expect given your article topic?
- **Interpretability**: read the top 10 terms at 2тАУ3 candidate k values тАФ do they form coherent, distinct themes?
- **LDAvis**: do the topic circles separate well, or do they overlap heavily?

Select a single best k and write 2тАУ3 sentences justifying your choice drawing on the three criteria.

**3. Fit your final model and visualize top terms**  
Plot the top 10 terms per topic.

**4. Visualize document-topic distributions**  
Plot ╬╕ for a sample of documents (5тАУ10). Choose a sample size that produces a readable plot.

**5. Interpret your topics**  
Assign a descriptive name to each topic (not just the concatenated top terms). In a short paragraph per topic, describe:
- What theme or issue does this topic represent?
- Which words are most diagnostic of this interpretation?
- Are there any surprising or ambiguous words in the top terms?

